Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-14 _tex/paper2.tex
Built site for gh-pages

It is easier to visualize the structure of these vectors and matrices by
considering an example. Assuming \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\), the vectors and matrices described in
equations (\ref{eq-mat11}) and (\ref{eq-mat12}) take the following form:
\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
}\end{equation}

Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{R}\), the texts' discriminal difference
\(D_{R}\), their discriminal processes \(T_{IA}\), and the judges'
biases \(B_{JK}\) are represented as vectors rather than scalar values.
These vectors capture all the observations from the conceptual
population. Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and
latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
\(B_{JK}\) are latent vectors of lengths \(n_{I} \cdot n_{A}\) and
\(n_{J} \cdot n_{K}\), respectively.

@@

Where each element of \(S_{I}\) is a binary value indicating the
presence or absence of corresponding elements in the vector \(I\), as in
equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\) using
vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and \(S_{J}\)
contains \(n_{I}\) and \(n_{J}\) elements, respectively.
\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is sampled} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
}\end{equation}

Similarly, each element of \(S_{IA}\) is a binary value indicating the
presence or absence of data rows in the matrices \(IA\), as defined in
equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\) using
the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
\(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
elements, respectively. \begin{equation}\phantomsection\label{eq-mat23}{
ia_{(i,a)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are sampled} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
}\end{equation}

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge, and they also omit the second text from
each individual and the third repeated judgment from each judge. Given
\(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
vectors take the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

@@

Finally, we define the \emph{sample mechanism} \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Each element \(s_{(i,a,h,b,j,k)}\) is a binary value
indicating the presence or absence of data rows in the matrix \(R\)
resulting from the sample mechanism, as in equation (\ref{eq-mat26}).
Thus, the vector contains \(n\) elements, matching the number of rows in
\(R\), and the sum of its elements represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model. Equation (\ref{eq-mat27})
shows that \(C\) contains \(n\) elements corresponding to the number of
rows in the \(R\) matrix, with each element \(c_{(i,a,h,b,j,k)}\) being
a binary value indicating the presence or absence of data rows in \(R\),
a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in equation
(\ref{eq-mat26}). \begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

@@

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
This derivation is possible because a fully specified SCM encodes
functional and probabilistic information, which we can replace with
suitable functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
to express the joint distribution of our complex CJ system as a product
of simpler conditional probability distributions (CPDs)\footnote{This
  re-expression is possible because the \emph{chain rule} of probability
  and the \emph{Bayesian Network Factorization (BNF)} property. For
  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
as shown in equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent. Here, \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}.
First, we assume that \(O_{R}\) follows a Bernoulli
distribution\footnote{The binomial distribution--including its special
  case, the Bernoulli distribution--represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that when only two un-ordered outcomes are possible, and the
  expected numbers of each type of event are assumed to be constant,
  then the binomial distribution is the one that is most consistent with
  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

\[
\begin{aligned}
  O_{R} & := f_{O}(D_{R}, S, C) \\ 
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
  e_{IA} & \:\bot\:e_{JK} \\ \\
\end{aligned}
\]

@@

\[
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\
  & P( e_{I} ) \\
  & P( e_{J} ) \\
  & P( e_{IA} ) \\  
  & P( e_{JK} )
\end{aligned}
\]

@@

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
\]

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
  an estimator---satisfies several key identification properties, such
  as common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
  spurious associations between the latter two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\),
representing the underlying written-quality trait of the texts, plus the
corresponding repeated judge biases, \(B_{JK}[j, k]\). Note that if we
assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
can re-write the discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). To clarify, the
square brackets in \(D_{R}\) simply indicate the relevant indices for
each trait vector and do not imply any subsetting of the data.

Third, we define \(T_{IA}\) as a linear combination of the students'
underlying writing-quality traits \(T_{I}\), the effects of relevant
text-related variables \(\beta_{XA}X_{IA}\) (e.g., text length on
quality), and the texts' idiosyncratic errors \(e_{IA}\). We assume that
each element of \(e_{IA}\) is independent and identically distributed
\((iid)\) according to a Normal distribution, with a mean of zero and a
standard deviation \(p_{IA}\). The parameter \(p_{IA}\) is defined as a
proportion of 1 to establish the scale of the latent trait, which is
required in any latent variable model \citep{Depaoli_2021}. Note that
because \(e_{IA}\) follows a Normal distribution, \(T_{IA}\) is also
normally distributed. This result implies that our model reverts to
assuming Normal discriminal processes for the stimuli, as originally
proposed by \citet{Thurstone_1927b} (see
Table~\ref{tbl-thurstone_cases}).

Fourth, we define \(T_{I}\) as a linear combination of the effects of
relevant student-related variables, \(\beta_{XI} X_{I}\), and students'
idiosyncratic errors, \(e_{I}\). As before, we assume that each element
of \(e_{I}\) is \(iid\) Normal with a mean of zero. However, we consider
the standard deviation, \(s_{XI}\), varies depending on the teaching
method group to which each student belongs. Given the example in
Section~\ref{sec-theoretical} where the teaching method
\(X_{I} = \{1,2\}\), the model sets the following constraint to anchor
the scale of the latent trait \citep{Depaoli_2021}:
\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\). Note that this parametrization of
\(e_{I}\) relaxes the equal dispersion assumption for the stimuli,
effectively addressing the concerns raised in
Section~\ref{sec-theory-issue1a}.

>>>>

It is easier to visualize the structure of these vectors and matrices by
considering an example. Assuming \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\), the vectors and matrices described in
Equations (\ref{eq-mat11}) and (\ref{eq-mat12}) take the following form:
\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
}\end{equation}

Now, using Equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{R}\), the texts' discriminal difference
\(D_{R}\), their discriminal processes \(T_{IA}\), and the judges'
biases \(B_{JK}\) are represented as vectors rather than scalar values.
These vectors capture all the observations from the conceptual
population. Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and
latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
\(B_{JK}\) are latent vectors of lengths \(n_{I} \cdot n_{A}\) and
\(n_{J} \cdot n_{K}\), respectively.

@@

Where each element of \(S_{I}\) is a binary value indicating the
presence or absence of corresponding elements in the vector \(I\), as in
Equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\) using
vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and \(S_{J}\)
contains \(n_{I}\) and \(n_{J}\) elements, respectively.
\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is sampled} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
}\end{equation}

Similarly, each element of \(S_{IA}\) is a binary value indicating the
presence or absence of data rows in the matrices \(IA\), as defined in
Equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\) using
the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
\(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
elements, respectively. \begin{equation}\phantomsection\label{eq-mat23}{
ia_{(i,a)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are sampled} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
}\end{equation}

We can visualize these vectors more clearly using the example in
Equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge, and they also omit the second text from
each individual and the third repeated judgment from each judge. Given
\(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
vectors take the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

@@

Finally, we define the \emph{sample mechanism} \(S\) in Equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Each element \(s_{(i,a,h,b,j,k)}\) is a binary value
indicating the presence or absence of data rows in the matrix \(R\)
resulting from the sample mechanism, as in Equation (\ref{eq-mat26}).
Thus, the vector contains \(n\) elements, matching the number of rows in
\(R\), and the sum of its elements represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model. Equation (\ref{eq-mat27})
shows that \(C\) contains \(n\) elements corresponding to the number of
rows in the \(R\) matrix, with each element \(c_{(i,a,h,b,j,k)}\) being
a binary value indicating the presence or absence of data rows in \(R\),
a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in Equation
(\ref{eq-mat26}). \begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

@@

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
This derivation is possible because a fully specified SCM encodes
functional and probabilistic information, which we can replace with
suitable functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
to express the joint distribution of our complex CJ system as a product
of simpler conditional probability distributions (CPDs)\footnote{This
  re-expression is possible because the \emph{chain rule} of probability
  and the \emph{Bayesian Network Factorization (BNF)} property. For
  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
as shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}. We
begin by assuming that \(O_{R}\) follows a Bernoulli
distribution\footnote{The binomial distribution--including its special
  case, the Bernoulli distribution--represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that the Bernoulli distribution is the most consistent
  alternative when only two un-ordered outcomes are possible and their
  expected frequencies are assumed to be constant
  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
  binomial as a maximum entropy distribution, see
  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
CJ outcomes. Furthermore, following the conventions of Generalized
Linear Models (GLMs)
\citep{Nelder_et_al_1983, Nelder_et_al_1996, Agresti_2015}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

\[
\begin{aligned}
  O_{R} & := f_{O}(D_{R}, S, C) \\ 
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\ \\
  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
  e_{IA} & \:\bot\:e_{JK}
\end{aligned}
\]

@@

\[
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\ \\
  & P( e_{I} ) P( e_{IA} ) P( e_{J} ) P( e_{JK} ) \\ \\ \\
\end{aligned}
\]

@@

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\ \\
  \boldsymbol{e} & \sim \text{Multi-Normal}( \boldsymbol{\mu}, \boldsymbol{\Sigma} )
  \\
  \boldsymbol{\Sigma} &= \boldsymbol{D} \boldsymbol{R} \boldsymbol{D} \\ \\
\end{aligned}
\]

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data--and, by extension,
  an estimator--satisfies several key identification properties, such as
  common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
  spurious associations between the latter two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Next we define \(D_{R}\) as the difference between the discriminal
processes \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), representing the
underlying written-quality trait of the texts, plus the corresponding
repeated judge biases \(B_{JK}[j, k]\). Note that if we assume that
\(B_{JK}[j,k]\) reflects the difference in stimulus-specific biases,
i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can
re-write the discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation}

This formulation reveals that the discriminal difference captures a
\emph{pure interaction effect}, in which neither the texts' discriminal
processes nor the judges' biases alone determine the outcome, but their
interaction does \citep{Attia_et_al_2022}. In simple terms, the
discriminal processes of the stimuli become an observable outcome only
through the judges' perceptions (biases). To clarify, the square
brackets in \(D_{R}\) indicate the relevant indices for each trait
vector; they do not imply any subsetting of the data.

We now specify the functional forms for \(T_{IA}\), \(T_{I}\),
\(B_{JK}\), and \(B_{J}\). We model \(T_{IA}\) as a linear combination
of the students' underlying writing-quality traits \(T_{I}\), the
effects of relevant text-related variables on quality assessment
\(\beta_{XA}X_{IA}\) (such as the influence of text length), and the
text-specific idiosyncratic errors \(e_{IA}\). Similarly, we express
\(T_{I}\) as a linear combination of relevant student-related variables
affecting the quality assessment \(\beta_{XI} X_{I}\), and
student-specific idiosyncratic errors \(e_{I}\). For the judge-specific
terms, we model \(B_{JK}\) as a linear combination of the judge's
individual bias \(B_{J}\), the influence of relevant judgment-related
variables on quality assessment \(\beta_{ZK}Z_{JK}\) (e.g., how the
number of judgments affect the evaluation), and judgment-specific
idiosyncratic errors \(e_{JK}\). Finally, we define \(B_{J}\) as a
linear combination of relevant judge-level variables influencing the
quality assessment \(\beta_{ZJ}Z_{J}\) (such as judgment expertise) and
judge-specific idiosyncratic errors \(e_{J}\).
